---
title: "Clustering and classification"
---

# Clustering and classification

*Tuomas Keski-Kuha 22.11.2021*

------------------------------------------------------

## Data exploration and finding correlations

----------------------------------------------------

*Exercise 2: Load the Boston data from the MASS package. Explore the structure and the dimensions of the data and describe the dataset briefly, assuming the reader has no previous knowledge of it. Details about the Boston dataset can be seen for example here.*

----------------------------------------------------

Let's see the Boston data in the following: 

```{r}
# access a few libraries: 
library(dplyr)
library(tidyr)
library(ggplot2)
library(corrplot)
library(GGally)
library(reshape2)

# access the MASS package
library(MASS)

# load the data
data("Boston")

# explore the dataset
str(Boston)
dim(Boston)

```

**Data set information**

Data is Housing Values in Suburbs of Boston. It has 506 rows and 14 columns. It has some information on the suburbs of Boston (towns).
Here is the [link](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html) to the data page, where one can see the full details. 

----------------------------------------------------

*Exercise 3: Show a graphical overview of the data and show summaries of the variables in the data. Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them.*

----------------------------------------------------

Let's visualize the data and take a look of the distributions of the variables and relations between variables. 

```{r}

summary(Boston)

# First lets make a long table from Boston data
melt.boston <- melt(Boston)

# Then plot every variable, so we'll see the distributions of variables
ggplot(data = melt.boston, aes(x = value)) + 
stat_density() + 
facet_wrap(~variable, scales = "free")

# calculate the correlation matrix and round it
cor_matrix<-cor(Boston) %>% round(digits = 2)

# visualize the correlation matrix
corrplot(cor_matrix, method="circle", type="upper", cl.pos="b", tl.pos="d", tl.cex = 0.6)


```


MUISTA TULKITA TULOKSET!!!!

MUISTA TULKITA TULOKSET!!!!

MUISTA TULKITA TULOKSET!!!!

MUISTA TULKITA TULOKSET!!!!

MUISTA TULKITA TULOKSET!!!!

MUISTA TULKITA TULOKSET!!!!



------------------------------------------------------

## Data preparing and making training and test datasets

----------------------------------------------------

*Exercise 4: Standardize the dataset and print out summaries of the scaled data. How did the variables change? Create a categorical variable of the crime rate in the Boston dataset (from the scaled crime rate). Use the quantiles as the break points in the categorical variable. Drop the old crime rate variable from the dataset. Divide the dataset to train and test sets, so that 80% of the data belongs to the train set.*

----------------------------------------------------

Let's scale the variables so that the mean = 0 and variance = 1 for all variables in the dataset: 

```{r}
# center and standardize variables
boston_scaled <- scale(Boston)

# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)

# summaries of the scaled variables
summary(boston_scaled)
 
```

We can see now that all the variables have negative and positive values around zero. 

In the following section we create a new categorical quantile variable from *crim* variable, so that the new variable has labels *low, med_low, med_high, high* depending the scaled *crim* variable, and all the new variables have 25 % of the data. 

```{r}

# create a quantile vector of crim and print it
bins <- quantile(boston_scaled$crim)

# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, labels = c("low", "med_low", "med_high", "high"))

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

```

After that we make training and test sets from the data, so that we pick randomly 80 % of the rows to the training set and the rest to the test set.

```{r}

# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

set.seed(123)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]

```

----------------------------------------------------

## Classification - Fitting the Linear Discriminant Analysis model

----------------------------------------------------

*Exercise 5: Fit the linear discriminant analysis on the train set. Use the categorical crime rate as the target variable and all the other variables in the dataset as predictor variables. Draw the LDA (bi)plot.*

----------------------------------------------------


```{r}

# linear discriminant analysis
lda.fit <- lda(crime ~ ., data = train)

# print the lda.fit object
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "orange", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)

```

----------------------------------------------------

*Exercise 6: Save the crime categories from the test set and then remove the categorical crime variable from the test dataset. Then predict the classes with the LDA model on the test data. Cross tabulate the results with the crime categories from the test set. Comment on the results. *

----------------------------------------------------

In the next phase we save crime gategories from the test set. Also we make predictions from test data, and see how accurate the model is: 

```{r}

# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)

# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)


```
Model's predicting power seems to be good indeed. Just few bigger inaccuracies, where correct class is two classes away. Still all results end up quite near the diagonal. Especially the high class seems to be "easy" to predict. 

----------------------------------------------------

## Clustering - Fitting the K-means model

----------------------------------------------------

*Exercise 7: Reload the Boston dataset and standardize the dataset (we did not do this in the Datacamp exercises, but you should scale the variables to get comparable distances). Calculate the distances between the observations. Run k-means algorithm on the dataset. Investigate what is the optimal number of clusters and run the algorithm again. Visualize the clusters (for example with the pairs() or ggpairs() functions, where the clusters are separated with colors) and interpret the results. *

----------------------------------------------------

```{r}
# Load the data again and after that standardize all variables (mean = 0 and variance = 1)
data("Boston")

# center and standardize variables
boston_scaled <- scale(Boston)

# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)

```

Then, we calculate the distances and apply K-means algorithm to make clusters:

```{r}

# euclidean distance matrix
dist_eu <- dist(boston_scaled)

# look at the summary of the distances
summary(dist_eu)

# k-means clustering
km <-kmeans(boston_scaled, centers = 3)

# plot the Boston scaled dataset with clusters
pairs(boston_scaled, col = km$cluster)

```

Finding the optimal value for the number of clusters: 

```{r}

set.seed(123)

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(Boston, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')

```

The optimal value is the value where the WCSS (within cluster sum of squares) drops radically (here the optimal number of clusters for K-means seems to be 2).

```{r}

# k-means clustering when k = 2
km <-kmeans(boston_scaled, centers = 2)

# plot the Boston scaled dataset with clusters
pairs(boston_scaled, col = km$cluster)

```

MUISTA TULKITA TULOKSET!!!!

MUISTA TULKITA TULOKSET!!!!

MUISTA TULKITA TULOKSET!!!!

MUISTA TULKITA TULOKSET!!!!

MUISTA TULKITA TULOKSET!!!!

MUISTA TULKITA TULOKSET!!!!

----------------------------------------------------

## Classification of clusters (Bonus)

----------------------------------------------------

*Bonus exercise: Perform k-means on the original Boston data with some reasonable number of clusters (> 2). Remember to standardize the dataset. Then perform LDA using the clusters as target classes. Include all the variables in the Boston data in the LDA model. Visualize the results with a biplot (include arrows representing the relationships of the original variables to the LDA solution). Interpret the results. Which variables are the most influencial linear separators for the clusters?*

----------------------------------------------------


```{r}
# Load the data again and after that standardize all variables (mean = 0 and variance = 1)
data("Boston")

# center and standardize variables
boston_scaled <- scale(Boston)

# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)

# k-means clustering
km <-kmeans(boston_scaled, centers = 5)

```

Let's add clusters to the data and then apply linear discrinmination analysis to the clusters that we got from K-means clustering: 

```{r}
# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, km$cluster)

set.seed(123)

# linear discriminant analysis
lda.fit <- lda(km.cluster ~ ., data = boston_scaled)

# print the lda.fit object
lda.fit

# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)

```

It seems clear that chas and rad variables are most influencial linear separators for the clusters. 
